---
title: "Stoops Effect"
author: Matthew Bladek
output:
  html_document:
    df_print: paged
---

###What is the Stroop Effect?

"In psychology, the Stroop effect is a demonstration of interference in the 
reaction time of a task.

When the name of a color (e.g., "blue", "green", or "red") is printed in a color
which is not denoted by the name (i.e., the word "red" printed in blue ink 
instead of red ink), naming the color of the word takes longer and is more prone 
to errors than when the color of the ink matches the name of the color."
<https://en.wikipedia.org/wiki/Stroop_effect>

### Stroop Task
"In a Stroop task, participants are presented with a list of words, with each 
word displayed in a color of ink. The participant’s task is to say out loud the 
color of the ink in which the word is printed. The task has two conditions: a congruent words condition, and an incongruent words condition. In the congruent 
words condition, the words being displayed are color words whose names match the colors in which they are printed: for example RED, BLUE. In the incongruent words
condition, the words displayed are color words whose names do not match the 
colors in which they are printed: for example PURPLE, ORANGE. In each case, we
measure the time it takes to name the ink colors in equally-sized lists. Each participant will go through and record a time from each condition."

<https://docs.google.com/document/d/1-OkpZLjG_kX9J6LIQ5IltsqMzVWjh36QpnP2RYpVdPU/pub?embedded=True>

### Experiment Design
The experiment was a website where it measured the time it took to name several 
words in both a base case and a treatment case.  The treatment being whether the 
words and the colors were congruent or not.

<https://faculty.washington.edu/chudler/java/ready.html>

In this experiment the dependent variable is the reaction time to name all of the
words on the screen.  We will refer to this as the "reaction time" from here on
out.  Our independent variable or treatment is whether the words are congruent
with their color or not.  

The null hypothesis is that there is no difference between reaction time or that
having incongruent colors actually decreases the reaction time.

The alternative hypothesis is that there is an increase in reaction time.  
We suspect that having incongruent colors would cause mental dissonance
and thus positively affect reaction time.  Thus we will perform a 1 sided t-test
in the positive direction.

The data provided was from Udacity, there were 24 observations giving the reaction
time before and after the treatment.  The time is in seconds.  I took the test
myself with a Congruent time of: 10.866 seconds and a Incongruent time of: 11.059.
We will addd this to our exisiting data to have a total of 25 observations

```{r echo = F, message= F, warning=F}
library(ggplot2)
library(dplyr)
setwd('~/__Education/R/Udacity_Data_Analysis/InferentialStats')
stroop <- read.csv('stroop.csv')
```



```{r echo = F, message= F, warning=F}
stroop[nrow(stroop) + 1, ] = list(10.886, 11.059)
stroop$diff <- stroop$Incongruent - stroop$Congruent
SDofDiff <- sd(stroop$diff)

mstroop <- subset(stroop, select = -c(diff))
mstroop$Type <- "Congruent"

# Restacked the data
for(row in 1:nrow(mstroop)){
  mstroop[nrow(mstroop) + 1, "Congruent"] <- mstroop[row, "Incongruent"] 
  mstroop[nrow(mstroop), "Type"] <- "Incongruent"
}
mstroop <- subset(mstroop, select = -c(Incongruent))

names(mstroop)[names(mstroop) == "Congruent"] <- "ReactionTime"

mstroop$Type <- factor(mstroop$Type)

```

### The Reaction Time Histrogram
```{r echo = F, message= F, warning=F}
ggplot(mstroop) +
  geom_freqpoly(aes(x = mstroop$ReactionTime, color = Type), binwidth = 1) +
  xlab("Reaction Time (sec)")
```

As we can see the distriubtion of the Congruent words and colors was centered
less than the Incongruent letters.  Both were roughly normally distributed.

There were two outliers above 30 sec for incongruent words.

### The Reaction Time BoxPlot
```{r echo = F, message= F, warning=F}
ggplot(mstroop, aes(x = Type, y = ReactionTime)) +
  geom_boxplot()
```

A box plot the two types shows a clear shift in the reaction time between the two.
We also see two outliers in the data >30 seconds which lie around 2 standard
deviations above the mean.  

We will remove these two cases from our study as they
are significant outliers.

```{r echo=FALSE}
remove_outlier <- function(df){
  for(row in 1:nrow(df)){
    if(df[row, "Incongruent"] > 30){
      df[row, "Incongruent"] = NA
    }
  }

  df <- df[complete.cases(df), ]
}
stroop <- remove_outlier(stroop)
```

### Experimental Results

Statistic  |Congruent (sec) | Incongruent (sec) | Differences (sec)
-- | -- | -- | --
Mean | 13.81 | 20.43 | 6.624 
SD  | | | 3.6077

Statistic | Value  
-- | --
Sample Size (N) |  23
Degrees of Freedom (DF)  | 22
Alpha Level | .05
One or Two sided | One sided, positive
tcritical | 1.7138
Average Mean Difference | 6.624 sec
Standard Eror of the Mean | 0.752 sec
Cohen's D | 1.835
R^2 | 77.8%
Margin of Error | 1.292 sec
Confidence Interval 95% Upper | 7.912 sec
Confidence Interval 95% Lower | 5.328 sec
t | 8.800
p-value | 5.852e-09

###Results:
CI = (5.328, 7.912)

t(22) = 8.800, p = 5.852e-09 < 0.05, one-tailed

The results are clear that there is an extremely low probability that reaction
time is decreased or not changed by the treatment.  Therefore, we reject the null hypothesis in favor of the alternative hypothesis that the
color congruency of the words does affect reaction time.

###Extensions
Exentsions of this project is to see what exactly is at the root of the difference.
One might hypothesize that it's the brains connection between color and writing
that has to be reprogrammed which casuses this lag.  Another study might involve
people who speak a foriegn language and give them this test.  There is no built
up association between the color and the word in the brain therefore one would
not expect to see that there would be a lag.  A longitudinal extension of this
would be to take this same group over a period of time, giving them the same test
and observing the differences once they start making the accociations between
the words and colors.

```{python echo = F, message= F, warning=F}
import math
import scipy.stats as st

# class for the alpha level
class AlphaT:
    def __init__(self, level, df): # level - the alpha level, df - the degrees of freedom
        
        self.level = level # alpha level as a porportion out of 1
        self.t1 = st.t.ppf(1 - level, df) # critical t-score for one tailed distribution
        self.t2 = st.t.ppf(1 - level/2, df) # critical t-score for two tailed distribution
   
# class for a t-test statistic
class StatsT:
    def __init__(self, mu, N, x, S):
        self.mu = mu # population mean
        self.N = N # sample size
        self.x = x # sample mean
        self.S = S # sample standard deviation
        self.SEM = S /math.sqrt(N) # sample t test
        self.df = N - 1 # degrees of freedom
        self.MD = x - mu # mean difference
        self.t = self.MD / (S / (math.sqrt(N)))
        self.cohenD = (self.x - self.mu) /self.S
        self.rsquared = math.pow(self.t, 2) / (math.pow(self.t,2) + self.df)
        
        self.a1 = AlphaT(.05, self.df) # alpha1 .05
        self.a2 = AlphaT(.01, self.df) # alpha2 .01
        self.a3 = AlphaT(.001, self.df) # alpha3 .001
    
    def getProb(self):
        return 1 - st.t.cdf(self.t, self.df) # calculates the probability from a given t-stat
    
    def getProbVal(self, tVal):
        return 1 - st.t.cdf(tVal, self.df)
    
    def getPVal(self, tVal, isTwoTailed):
        if(isTwoTailed):
            return 2 *(1 - st.t.cdf(tVal, self.df))
        else:
            return (1 - st.t.cdf(tVal, self.df))
            
stroopT = StatsT(13.81, 23, 20.43, 3.6077)

#print(stroopT.a1.t1)
#print(stroopT.SEM)
#print(stroopT.SEM * stroopT.a1.t1)
#print(stroopT.t)
#print(stroopT.cohenD)
#print(stroopT.rsquared)
#print(stroopT.getPVal(stroopT.t, False))
#print(stroopT.MD + stroopT.SEM * stroopT.a1.t1)

```
